diffEnrich: An \proglang{R} Package to Compare Functional Enrichment Between Two Experimentally-derived Groups of Genes by Connecting to the KEGG REST API

### 2019-12-05 jss-article-rnw-diffEnrich/article.Rnw
incorp supp info into text body and remove supp section

By default, p-values from the Fisher’s Exact test are adjusted for multiple
comparisons with a False Discovery Rate (FDR) (Benjamini:1995), however users
have the option of choosing any type of multiple testing correction supported
by \code{p.adjust}. In addition to the unadjusted p-value and FDR, pathEnrich
will calculate for each KEGG pathway, its fold enrichment defined as the ratio
of number of genes observed from the gene list annotated to the KEGG pathway to
the expected number of genes from the gene list to be annotated to the KEGG
pathway by chance. An example of the first 6 results generated by \code{pathEnrich}
are in Supplementary Table 1. For a detailed description of the variables in this table see Table 2.

@@

S3 generic functions for \code{print} and \code{summary} are provided.
The \code{print} function prints the results table as a \code{tibble}, and the
\code{summary} function returns the number of pathways that reached statistical
significance as well as their descriptions, the number of genes used from the
KEGG data base, the KEGG species, and the method used for multiple testing
correction. See supplementary information for examples.

@@

The \code{diffEnrich} function merges results from the \code{pathEnrich} function
generated in section 2.2. This merged data set is then used to perform differential
enrichment using a Fisher’s exact test as described in 2.2. The resulting odds
ratio is defined as the odds of a gene from list 2 belonging to a given KEGG
pathway divided by the odds of a gene from list 1 belonging to a given KEGG
pathway. Users have the same options for multiple testing methods that are
provided in the \code{pathEnrich} function. KEGG pathways that do not contain
any genes from either gene list (e.g. ‘rno04530’ contains 0 genes from list 1
and 0 genes from list 2) are removed from the analysis. If this is the case a
warning will be printed that tells the user how many pathways were removed.
This can be avoided by setting the 'N' parameter to a value > 0 in the
\code{pathEnrich calls}. This diffEnrich function generates a table that
contains the results from the analyses performed in section 2.2 for each gene
list as well as odds ratios and their associated unadjusted and adjusted
p-values for each KEGG pathway. For a detailed description of the results
generated by \code{diffEnrich} see Table 3, and an example of the first 6
results generated by \code{diffEnrich} are in Supplementary Table 2.

@@

@@

@@

%% -----Supplementary Information------------------------------------
\section[Supplementary Information]{Supplementary Information} \label{sec:supinfo}

@@

\begin{table}[H]
  \centering
  \caption{Supplementary Table 1. First 6 results generated by pathEnrich.}
  \begin{adjustbox}{width=\textwidth, angle=90}
  \fontsize{90}{125}\selectfont
  \begin{tabular}{rllrrrrrrrr}
  \hline
 & KEGG\_PATHWAY\_ID & KEGG\_PATHWAY\_description & KEGG\_PATHWAY\_cnt & KEGG\_PATHWAY\_in\_list & KEGG\_DATABASE\_cnt & KEG\_DATABASE\_in\_list & expected & enrich\_p & p\_adj & fold\_enrichment \\
  \hline
95 & rno04530 & Tight junction - Rattus norvegicus (rat) & 170 & 19.00 & 8856 & 295 & 5.66 & 0.00 & 0.00 & 3.36 \\
  172 & rno05135 & Yersinia infection - Rattus norvegicus (rat) & 128 & 16.00 & 8856 & 295 & 4.26 & 0.00 & 0.00 & 3.75 \\
  194 & rno05210 & Colorectal cancer - Rattus norvegicus (rat) &  88 & 12.00 & 8856 & 295 & 2.93 & 0.00 & 0.00 & 4.09 \\
  212 & rno05231 & Choline metabolism in cancer - Rattus norvegicus (rat) &  99 & 12.00 & 8856 & 295 & 3.30 & 0.00 & 0.00 & 3.64 \\
  197 & rno05213 & Endometrial cancer - Rattus norvegicus (rat) &  58 & 9.00 & 8856 & 295 & 1.93 & 0.00 & 0.00 & 4.66 \\
  66 & rno04144 & Endocytosis - Rattus norvegicus (rat) & 275 & 22.00 & 8856 & 295 & 9.16 & 0.00 & 0.00 & 2.40 \\
   \hline
  \end{tabular}
\end{adjustbox}
\end{table}

@@

\begin{table}[H]
\centering
  \caption{Supplementary Table 2. First 6 results generated by diffEnrich}
  \begin{adjustbox}{width=\textwidth, angle=90}
  \fontsize{50}{80}\selectfont
\begin{tabular}{rllrrrrrrrrrrrrrrrrr}
  \hline
 & KEGG\_PATHWAY\_ID & KEGG\_PATHWAY\_description & KEGG\_PATHWAY\_cnt & KEGG\_DATABASE\_cnt & KEGG\_PATHWAY\_in\_list1 & KEGG\_DATABASE\_in\_list1 & expected\_list1 & enrich\_p\_list1 & p\_adj\_list1 & fold\_enrichment\_list1 & KEGG\_PATHWAY\_in\_list2 & KEGG\_DATABASE\_in\_list2 & expected\_list2 & enrich\_p\_list2 & p\_adj\_list2 & fold\_enrichment\_list2 & odd\_ratio & diff\_enrich\_p & diff\_enrich\_adjusted \\
  \hline
rno04530 & rno04530 & Tight junction - Rattus norvegicus (rat) & 170 & 8856 & 19.00 & 295 & 5.66 & 0.00 & 0.00 & 3.36 & 131.00 & 5308 & 101.89 & 0.00 & 0.00 & 1.29 & 0.37 & 0.00 & 0.00 \\
  rno05135 & rno05135 & Yersinia infection - Rattus norvegicus (rat) & 128 & 8856 & 16.00 & 295 & 4.26 & 0.00 & 0.00 & 3.75 & 105.00 & 5308 & 76.72 & 0.00 & 0.00 & 1.37 & 0.35 & 0.00 & 0.00 \\
  rno05210 & rno05210 & Colorectal cancer - Rattus norvegicus (rat) &  88 & 8856 & 12.00 & 295 & 2.93 & 0.00 & 0.00 & 4.09 & 81.00 & 5308 & 52.74 & 0.00 & 0.00 & 1.54 & 0.37 & 0.00 & 0.00 \\
  rno05213 & rno05213 & Endometrial cancer - Rattus norvegicus (rat) &  58 & 8856 & 9.00 & 295 & 1.93 & 0.00 & 0.00 & 4.66 & 55.00 & 5308 & 34.76 & 0.00 & 0.00 & 1.58 & 0.33 & 0.01 & 0.01 \\
  rno04660 & rno04660 & T cell receptor signaling pathway - Rattus norvegicus (rat) & 106 & 8856 & 11.00 & 295 & 3.53 & 0.00 & 0.01 & 3.12 & 79.00 & 5308 & 63.53 & 0.00 & 0.00 & 1.24 & 0.39 & 0.01 & 0.01 \\
  rno04657 & rno04657 & IL-17 signaling pathway - Rattus norvegicus (rat) &  95 & 8856 & 9.00 & 295 & 3.16 & 0.00 & 0.03 & 2.84 & 59.00 & 5308 & 56.94 & 0.37 & 0.46 & 1.04 & 0.36 & 0.01 & 0.01 \\
   \hline
\end{tabular}
\end{adjustbox}
\end{table}

Example of S3 functions
<<path_summary, echo = TRUE>>=
summary(list1_pe)
@

>>>>

By default, p-values from the Fisher’s Exact test are adjusted for multiple
comparisons with a False Discovery Rate (FDR) (Benjamini:1995), however users
have the option of choosing any type of multiple testing correction supported
by \code{p.adjust}. In addition to the unadjusted p-value and FDR, pathEnrich
will calculate for each KEGG pathway, its fold enrichment defined as the ratio
of number of genes observed from the gene list annotated to the KEGG pathway to
the expected number of genes from the gene list to be annotated to the KEGG
pathway by chance. An example of the first 6 results generated by \code{pathEnrich}
are in displayed below. For a detailed description of the variables in this table see Table 2.

<<pathEnrich_example, echo=TRUE>>=
head(list1_pe$enrich_table)
@

@@

S3 generic functions for \code{print} and \code{summary} are provided.
The \code{print} function prints the results table as a \code{tibble}, and the
\code{summary} function returns the number of pathways that reached statistical
significance as well as their descriptions, the number of genes used from the
KEGG data base, the KEGG species, and the method used for multiple testing
correction.

<<path_summary, echo = TRUE>>=
summary(list1_pe)
@

@@

The \code{diffEnrich} function merges results from the \code{pathEnrich} function
generated in section 2.2. This merged data set is then used to perform differential
enrichment using a Fisher’s exact test as described in 2.2. The resulting odds
ratio is defined as the odds of a gene from list 2 belonging to a given KEGG
pathway divided by the odds of a gene from list 1 belonging to a given KEGG
pathway. Users have the same options for multiple testing methods that are
provided in the \code{pathEnrich} function. KEGG pathways that do not contain
any genes from either gene list (e.g. ‘rno04530’ contains 0 genes from list 1
and 0 genes from list 2) are removed from the analysis. If this is the case a
warning will be printed that tells the user how many pathways were removed.
This can be avoided by setting the 'N' parameter to a value > 0 in the
\code{pathEnrich calls}. This diffEnrich function generates a table that
contains the results from the analyses performed in section 2.2 for each gene
list as well as odds ratios and their associated unadjusted and adjusted
p-values for each KEGG pathway. For a detailed description of the results
generated by \code{diffEnrich} see Table 3, and an example of the first 6
results generated by \code{diffEnrich} are displayed below.

@@

<<diffEnrich_example, echo=TRUE>>=
head(diff_enrich$de_table)
@

@@

<<diff_summary, echo = TRUE>>=
summary(diff_enrich)
@

@@

%% -----Supplementary Information------------------------------------
% \section[Supplementary Information]{Supplementary Information} \label{sec:supinfo}

@@

% \begin{table}[H]
%   \centering
%   \caption{Supplementary Table 1. First 6 results generated by pathEnrich.}
%   \begin{adjustbox}{width=\textwidth, angle=90}
%   \fontsize{90}{125}\selectfont
%   \begin{tabular}{rllrrrrrrrr}
%   \hline
%  & KEGG\_PATHWAY\_ID & KEGG\_PATHWAY\_description & KEGG\_PATHWAY\_cnt & KEGG\_PATHWAY\_in\_list & KEGG\_DATABASE\_cnt & KEG\_DATABASE\_in\_list & expected & enrich\_p & p\_adj & fold\_enrichment \\
%   \hline
% 95 & rno04530 & Tight junction - Rattus norvegicus (rat) & 170 & 19.00 & 8856 & 295 & 5.66 & 0.00 & 0.00 & 3.36 \\
%   172 & rno05135 & Yersinia infection - Rattus norvegicus (rat) & 128 & 16.00 & 8856 & 295 & 4.26 & 0.00 & 0.00 & 3.75 \\
%   194 & rno05210 & Colorectal cancer - Rattus norvegicus (rat) &  88 & 12.00 & 8856 & 295 & 2.93 & 0.00 & 0.00 & 4.09 \\
%   212 & rno05231 & Choline metabolism in cancer - Rattus norvegicus (rat) &  99 & 12.00 & 8856 & 295 & 3.30 & 0.00 & 0.00 & 3.64 \\
%   197 & rno05213 & Endometrial cancer - Rattus norvegicus (rat) &  58 & 9.00 & 8856 & 295 & 1.93 & 0.00 & 0.00 & 4.66 \\
%   66 & rno04144 & Endocytosis - Rattus norvegicus (rat) & 275 & 22.00 & 8856 & 295 & 9.16 & 0.00 & 0.00 & 2.40 \\
%    \hline
%   \end{tabular}
% \end{adjustbox}
% \end{table}

@@

% \begin{table}[H]
% \centering
%   \caption{Supplementary Table 2. First 6 results generated by diffEnrich}
%   \begin{adjustbox}{width=\textwidth, angle=90}
%   \fontsize{50}{80}\selectfont
% \begin{tabular}{rllrrrrrrrrrrrrrrrrr}
%   \hline
%  & KEGG\_PATHWAY\_ID & KEGG\_PATHWAY\_description & KEGG\_PATHWAY\_cnt & KEGG\_DATABASE\_cnt & KEGG\_PATHWAY\_in\_list1 & KEGG\_DATABASE\_in\_list1 & expected\_list1 & enrich\_p\_list1 & p\_adj\_list1 & fold\_enrichment\_list1 & KEGG\_PATHWAY\_in\_list2 & KEGG\_DATABASE\_in\_list2 & expected\_list2 & enrich\_p\_list2 & p\_adj\_list2 & fold\_enrichment\_list2 & odd\_ratio & diff\_enrich\_p & diff\_enrich\_adjusted \\
%   \hline
% rno04530 & rno04530 & Tight junction - Rattus norvegicus (rat) & 170 & 8856 & 19.00 & 295 & 5.66 & 0.00 & 0.00 & 3.36 & 131.00 & 5308 & 101.89 & 0.00 & 0.00 & 1.29 & 0.37 & 0.00 & 0.00 \\
%   rno05135 & rno05135 & Yersinia infection - Rattus norvegicus (rat) & 128 & 8856 & 16.00 & 295 & 4.26 & 0.00 & 0.00 & 3.75 & 105.00 & 5308 & 76.72 & 0.00 & 0.00 & 1.37 & 0.35 & 0.00 & 0.00 \\
%   rno05210 & rno05210 & Colorectal cancer - Rattus norvegicus (rat) &  88 & 8856 & 12.00 & 295 & 2.93 & 0.00 & 0.00 & 4.09 & 81.00 & 5308 & 52.74 & 0.00 & 0.00 & 1.54 & 0.37 & 0.00 & 0.00 \\
%   rno05213 & rno05213 & Endometrial cancer - Rattus norvegicus (rat) &  58 & 8856 & 9.00 & 295 & 1.93 & 0.00 & 0.00 & 4.66 & 55.00 & 5308 & 34.76 & 0.00 & 0.00 & 1.58 & 0.33 & 0.01 & 0.01 \\
%   rno04660 & rno04660 & T cell receptor signaling pathway - Rattus norvegicus (rat) & 106 & 8856 & 11.00 & 295 & 3.53 & 0.00 & 0.01 & 3.12 & 79.00 & 5308 & 63.53 & 0.00 & 0.00 & 1.24 & 0.39 & 0.01 & 0.01 \\
%   rno04657 & rno04657 & IL-17 signaling pathway - Rattus norvegicus (rat) &  95 & 8856 & 9.00 & 295 & 3.16 & 0.00 & 0.03 & 2.84 & 59.00 & 5308 & 56.94 & 0.37 & 0.46 & 1.04 & 0.36 & 0.01 & 0.01 \\
%    \hline
% \end{tabular}
% \end{adjustbox}
% \end{table}
